Distinguishing Between Spin-Aligned and Isotropic Binary Black Hole Populations Using Gravitational Wave Observations

### 2017-03-27 paper/AlignedIsotropicGW.tex
Some text polishing and discussion of difference between measuring angular and magnitude distribution

\newcommand{\OOneSigmaIsoAligned}{\checkme{1.8}}
\newcommand{\OOneOddsIsoAligned}{\checkme{0.08}}
\newcommand{\OTwoSigmaIsoAlignedMin}{\checkme{3.0}}
\newcommand{\OTwoOddsIsoAlignedMin}{\checkme{0.0012}}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\citep{2015PhRvD..91d2003V} compare model gravitational waveforms
\citep{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameter space that describes the compact binary source.  The spin
parameter with the largest effect on waveforms, and a correspondingly
tight constraint from the data, is a mass-weighted, aligned,
``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\hat{L}} = \frac{1}{M} \left( m_1 a_1^z + m_2 a_2^z \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $0 \leq a_{1,2} \leq 1$ are the corresponding dimensionless spin
parameters \citep{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  There is essentially no posterior support for
$\chieff \gtrsim 0.5$ as observed in the majority of the
electromagnetically-detected stellar-mass black hole population.  The
hierarchical modelling is relatively insensitive to the precise
details of the posterior distributions, unlike single-source analysis;
notably our approximation does permit $\chieff = 0$ for GW151226 while
the true posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

@@

Small values of $\chieff$ can result from either intrinsically small
spins or larger spins whose direction is mis-aligned with the orbital
plane of the binary (i.e.\ spin vectors with small $z$-components).
Mis-alignment is capable of producing \emph{negative} values of
$\chieff$, however, while small, aligned spins will always have
$\chieff \geq 0$.  To quantify the degree of support for these two
alternate explanations of small $\chieff$ values in the merging
\ac{BBH} population, we compared the Bayesian evidence for various
simple models of the spin population using the \ac{GW} data set.

@@

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``decreasing'' magnitude distribution; the most
favoured model among those with an aligned distribution also has the
``decreasing'' magnitude distribution.  The odds ratio between the
aligned and isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the the correct angular distribution to
be distinguished with greater than $5\sigma$ ($2.9 \times 10^{-7}$
odds) credibility; and in the worst case the distinction is
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).

\begin{figure}
  \plotone{../plots/six-way-O2-model-selection}
  \caption{\label{fig:O2-predictions} Distribution of odds ratios
    predicted with 10 additional observations above the three
    discussed in Section \ref{sec:O1}.  Each panel corresponds to
    additional observations drawn from one of the $\chieff$
    distribution models.  The model from which the additional
    observations are drawn is outlined in red.  The height of the blue
    bar gives the median odds ratio relative to the model from which
    the additional observations are drawn; the green line gives the
    68\% ($1 \sigma$) symmetric interval of odds ratios over 1000
    separate draws from the model distribution.  The closest ratio
    between the most-favoured isotropic model and the most-favoured
    aligned model is $\OTwoOddsIsoAlignedMin$, corresponding to
    $\OTwoSigmaIsoAlignedMin\sigma$ preference for the correct angular
    distribution; most models result in more than $5\sigma$ preference
    for the correct angular distribution.  Because the three
    observations from Section \ref{sec:O1} are included in each data
    set the ``correct'' model is not necessarily preferred over the
    others, particularly when that model uses the ``high'' magnitude
    distribution, which is strongly dis-favoured from the O1
    observations alone.}
\end{figure}

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.1$ always, then
it is no longer possible to resolve the negative spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large, and most models of isolated binary
evolution do not produce systems with spins uniformly bounded to small
values.

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate the models.  The ``aligned'' population in that
study permitted spins nearly parallel or anti-parallel to the orbital
axis, i.e.\ $\hat{S}\cdot \hat{L} \simeq \pm 1$, in contrast to our
aligned models.  Anti-parallel spins are strongly disfavoured in
models of isolated binary evolution \citep[e.g.,]{Kalogera:2000}.

In order to perform our analysis we need to select models for the
distribution of the dimensionless spin parameters of stellar-mass
black holes.  Data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.

Most of the systems studied are low-mass X-ray binaries (i.e., the
donor star has a mass $M_{\rm donor}<1~M_\odot$) rather than the
high-mass X-ray binaries that are likely to be the progenitors of
double black hole binaries.  In addition, there are substantial
systematic errors that can complicate either type of analysis (see
\citealt{2015PhR...548....1M} for a discussion). Nonetheless, if we
take the reported spin magnitudes as representative then we find that
there is a preference for high spins; for example, 14 of the 19
systems with reported spins have spin parameters in excess of 0.5.
The masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors
simply do not have enough mass and high-mass donors have a very short
phase in which they transfer mass; a variant of this long-standing
argument was presented by \citet{1999MNRAS.305..654K}).  Thus the
current spin parameters probably are close to their values upon core
collapse.  We emphasize, however, that the specific processes involved
in the production of black hole binaries could alter the spin
distribution of those holes: for example, close tidal interactions
could spin up the core, or stripping of the envelope could reduce the
available angular momentum.

@@

There are a few modelling constrains on the predicted spin
distribution.  We separate these into a discussion of spin magnitudes
and spin directions (spin-orbit misalignment angles) as in the
preceding analysis; however, it is worth pointing out that, contrary
to our simplified assumptions, the magnitude and direction
distributions may well be coupled.

Several models for binary black hole formation have been proposed
\citep[e.g.,][]{2016ApJ...818L..22A}, yielding potentially different spin
distributions.

@@

In short, the angular distribution of spins in \acp{BBH} formed
through isolated binary evolution is uncertain; nevertheless, it is
more probable that spins are aligned and non-zero after evolution
through this channel than not.

>>>>

\newcommand{\OOneSigmaIsoAligned}{\checkme{1.8}}
\newcommand{\OOneOddsIsoAligned}{\checkme{0.08}}
\newcommand{\OTwoSigmaIsoAlignedMin}{\checkme{2.9}}
\newcommand{\OTwoOddsIsoAlignedMin}{\checkme{0.0035}}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\citep{2015PhRvD..91d2003V} compare model gravitational waveforms
\citep{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameter space that describes the compact binary source.  The model
waveforms incorporate approximations to the true behaviour of the
merging systems that are expected to break down for sufficiently high
mis-aligned spins; detailed comparisons with numerical relativity
computations using no approximations to the dynamics
\citep{2016PhRvD..94f4035A} show no evidence of such a breakdown in
GW150914's analysis.  Given the preference of our data for smaller
spins (discussed below) we assume here that measurements made now and
in the future are not dominated by systematic errors.

The spin parameter with the largest effect on waveforms, and a
correspondingly tight constraint from the data, is a mass-weighted,
aligned, ``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\hat{L}} = \frac{1}{M} \left( m_1 a_1^z + m_2 a_2^z \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $0 \leq a_{1,2} \leq 1$ are the corresponding dimensionless spin
parameters \citep{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  There is essentially no posterior support for
$\chieff \gtrsim 0.5$ as observed in the majority of the
electromagnetically-detected stellar-mass black hole population (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

@@

Small values of $\chieff$ can result from either intrinsically small
spins or larger spins whose direction is mis-aligned with the orbital
plane of the binary (i.e.\ spin vectors with small $z$-components).
Mis-alignment is capable of producing \emph{negative} values of
$\chieff$, however, while small, aligned spins will always have
$\chieff \geq 0$.  This difference provides strong discriminating
power between the two angular distributions, even without good
information about the magnitude distribution; to the extent that data
support negative $\chieff$ they weigh heavily against aligned models.
To quantify the degree of support for these two alternate explanations
of small $\chieff$ values in the merging \ac{BBH} population, we
compared the Bayesian evidence for various simple models of the spin
population using the \ac{GW} data set.

@@

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``decreasing'' magnitude distribution; the most
favoured model among those with an aligned distribution also has the
``decreasing'' magnitude distribution.  The odds ratio between the
aligned and isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound $\chieff \gtrsim 0.2$ at 90\% credibility
\citep{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the the correct angular distribution to
be distinguished with greater than $5\sigma$ ($2.9 \times 10^{-7}$
odds) credibility; and in the worst case the distinction is
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
Note that such future detections should permit a confident distinction
between \emph{angular} distributions, we will remain very uncertain
about the \emph{magnitude} distribution until we have a much larger
number of observations.  In Figure \ref{fig:O2-predictions}, the odds
ratio between different magnitude distributions with the same angular
distribution is much closer to one than the odds ratio between angular
distributions.

\begin{figure}
  \plotone{../plots/six-way-O2-model-selection}
  \caption{\label{fig:O2-predictions} Distribution of odds ratios
    predicted with 10 additional observations above the three
    discussed in Section \ref{sec:O1}.  Each panel corresponds to
    additional observations drawn from one of the $\chieff$
    distribution models.  The model from which the additional
    observations are drawn is outlined in red.  The height of the blue
    bar gives the median odds ratio relative to the model from which
    the additional observations are drawn; the green line gives the
    68\% ($1 \sigma$) symmetric interval of odds ratios over 1000
    separate draws from the model distribution.  The closest median
    ratio between the most-favoured isotropic model and the
    most-favoured aligned model is $\OTwoOddsIsoAlignedMin$,
    corresponding to $\OTwoSigmaIsoAlignedMin\sigma$ preference for
    the correct angular distribution; most models result in more than
    $5\sigma$ preference for the correct angular distribution.
    Because the three observations from Section \ref{sec:O1} are
    included in each data set the ``correct'' model is not necessarily
    preferred over the others, particularly when that model uses the
    ``high'' magnitude distribution, which is strongly dis-favoured
    from the O1 observations alone.}
\end{figure}

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.1$ always, then
it is no longer possible to resolve the negative spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large, and most models of isolated binary
evolution do not produce systems with spins uniformly bounded to small
values.  Additionally, models which do not permit \emph{some} spins
with $\chieff \gtrsim 0.2$ are ruled out by the GW151226 observations
\citep{2016PhRvL.116x1103A}.

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate the models.  The ``aligned'' population in that
study permitted spins nearly parallel or anti-parallel to the orbital
axis, i.e.\ $\hat{S}\cdot \hat{L} \simeq \pm 1$, in contrast to our
aligned models; thus the aligned/isotropic discrimitating power in
\citet{2017CQGra..34cLT01V} relies effectively on measurements of the
detailed shape of the $\chieff$ distribution and, as noted above,
measurements are not very sensitive to the spin magnitude distribution
that controls the shape of $\chieff$.  Note that anti-parallel spins
are strongly disfavoured in models of isolated binary evolution (see,
for example, \citet{Kalogera:2000}).

In order to perform our analysis we need to select models for the
distribution of the dimensionless spin parameters of stellar-mass
black holes.  Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries (i.e., the donor star has a mass
$M_{\rm donor}<1~M_\odot$) rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion). Nonetheless, if we take the reported spin
magnitudes as representative then we find that there is a preference
for high spins; for example, 14 of the 19 systems with reported spins
have spin parameters in excess of 0.5.  The masses and spin parameters
of stellar-mass black holes are unlikely to be altered significantly
by accretion (low-mass donors simply do not have enough mass and
high-mass donors have a very short phase in which they transfer mass;
a variant of this long-standing argument was presented by
\citet{1999MNRAS.305..654K}).  Thus the current spin parameters
probably are close to their values upon core collapse; however, the
specific processes involved in the production of black hole binaries
could alter the spin distribution of those holes.  For example, close
tidal interactions could spin up the core, or stripping of the
envelope could reduce the available angular momentum.

@@

There are a few modelling constrains on the predicted spin
distribution (see, for example, \citep{2016ApJ...818L..22A}).  We
separate these into a discussion of spin magnitudes and spin
directions (spin-orbit misalignment angles) as in the preceding
analysis; however, it is worth pointing out that, contrary to our
simplified assumptions, the magnitude and direction distributions may
well be coupled.

@@

In short, the angular distribution of spins in \acp{BBH} formed
through isolated binary evolution is uncertain; nevertheless, it is
more probable that spins are aligned and non-zero after evolution
through this channel than not, while an isotropic distribution of
non-zero spins is the natural outcome of dynamical formation
processes.

\acknowledgements

We thank Richard O’Shaughnessy for comments on this work.
